---
title: "Voting Patterns Unveiled"
subtitle: "The Impact of Demographics and Economic Outlook in the 2020 Election"
author: 
  - Raghav Bhatia
thanks: "Code and data are available at: https://github.com/1raghav-bhatia/Spectral-Models.git"
date: April 16, 2024
date-format: long
abstract: "This paper investigates voter demographics and economic outlook on preferences for candidates in the 2020 US presidential election. It was found that race and education played pivotal roles in determining support for Biden over Trump. Moreover, economic factors influenced voter sentiment with perceptions of economic hardship impacting candidate favorability. Awareness of these complex dynamics is imperative for grasping the landscape of American politics especially with the upcoming US election this November."
toc: true
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(knitr)
library(modelsummary)
library(rstanarm)
library(tidyverse)
library(readr)
library(arrow)
library(knitr)
library(quantmod)
library(forecast)
library(wavethresh)
library(kableExtra)

### Reading the raw data ###

sp500_raw <- arrow::read_parquet("data/raw_data/sp500_data.parquet")
vix_raw <- arrow::read_parquet("data/raw_data/vix_data.parquet")

### Reading the cleaned Data ###


### Reading the Model ###

```


# Introduction

In an era shaped by rapid shifts in market sentiment and extensive global information flows, understanding the dynamics between market shocks and future short-term market volatility expectations remains crucial for both investors and policymakers. This research investigates the predictive relationship between short-term market shocks, as measured by the residuals of market returns, and their impact on future short-term market volatility expectations, represented by the Volatility Index (VIX). By examining the lagged effects of these shocks, this study challenges the efficient markets hypothesis [@fama1970efficient], which asserts that current prices fully reflect all available information and thus, no amount of historical data can predict future movements.

The premise of this investigation is rooted in the observation that market participants often exhibit behavioral biases, such as the momentum effect [@jegadeesh1993returns], where investors assume that current market conditions will persist into the future. This bias suggests that positive market returns, which indicate good current conditions, might lead to a decrease in expected future risk, as reflected by the VIX. Conversely, negative shocks might increase future risk aversion, leading to higher implied volatility. These behaviors challenge the expectations of the efficient markets hypothesis, indicating a potential disconnect between market participant behavior and market efficiency theories.

To explore this relationship, this study employs a generalized linear model (GLM) using detailed coefficients derived from Discrete Haar Transforms [@percival2000wavelet] of both market returns and VIX data. The primary estimand in this study is the effect of short-term market return shocks on the future short-term volatility expectations as measured by the VIX, particularly focusing on a two-month lag. This approach not only allows for the isolation of short-term fluctuations but also aligns with sophisticated financial modeling techniques that account for non-linear dependencies and temporal dynamics in financial data.

This paper is structured to first detail the methods used for data collection and transformation, specifically focusing on the use of wavelet analysis to extract meaningful patterns from high-frequency financial data. Additionally, we construct a GLM model to showcase our methodology for analyzing this relationship. Following this, we present the results of our regression analysis, highlighting the significant predictive power of market shocks on future volatility. We conclude with a discussion on the implications of our findings for financial theory and practice, emphasizing how traditional models might need to adapt to incorporate behavioral biases and their effects on market dynamics.

By bridging the gap between theoretical finance and empirical evidence, this research contributes to a deeper understanding of market dynamics and offers valuable insights into the predictive relationships that might inform future trading strategies and risk management practices.


# Data {#sec-data}

## Data Source and Preparation

This research utilizes publicly available financial data for the Standard & Poor's 500 Index (S&P 500) and the Chicago Board Options Exchange Volatility Index (VIX), sourced from Yahoo Finance. The choice of Yahoo Finance as a data source is driven by its comprehensive coverage and reliable historical financial data accessibility, which is crucial for the integrity of time-series analysis in financial research. The data covers a period from January 1993 to December 2023 for the S&P 500 and from March 1993 to February 2024 for the VIX.

The specific variables extracted from these datasets are the daily closing prices, which are essential for calculating monthly returns and analyzing market behavior over the study period. This selection is informed by the closing price's relevance in reflecting the final market consensus on value for each day, making it a critical indicator for financial analysis.

While alternative sources such as Google Finance or Bloomberg could provide similar data, Yahoo Finance was selected due to its ease of integration with the quantmod [@quantmod] package in R, which significantly simplifies the process of data fetching and preliminary handling. Furthermore, Yahoo Finance offers unrestricted access to historical data, unlike some other services that might require subscriptions.

The data retrieval and initial processing were conducted using the quantmod package in R, which is instrumental for financial modeling and time series analysis due to its capabilities in fetching and handling various financial data formats.

## Data Cleaning and Transformation

Once retrieved, the data was cleaned and transformed into monthly returns to average out the minor fluctuations in day-to-day returns and to capture the significant monthly trends that highlight short-term market fluctuations. This transformation was facilitated by converting daily data into monthly data points, specifically extracting closing prices and calculating the monthly returns. The transformation process employed the xts [@xts] library in R.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-monthly-return-vix

library(knitr)
library(kableExtra)

### Transforming Data to Monthly Returns and Cleaning
# S&P 500
sp500_monthly <- to.monthly(sp500_raw, indexAt = "lastof", OHLC = FALSE)
sp500_returns <- ROC(Cl(sp500_monthly)) * 100  # Convert to percentage returns
sp500_clean <- na.omit(sp500_returns)  # Remove NAs

# VIX
vix_monthly <- to.monthly(vix_raw, indexAt = "lastof", OHLC = FALSE)
vix_clean <- ROC(Cl(vix_monthly)) * 100
vix_clean <- na.omit(vix_clean)

# Assuming sp500_clean and vix_clean are the cleaned monthly returns data

# Combine the two series into one data frame
combined_data <- data.frame(Date = index(sp500_clean), 
                            SP500_Returns = coredata(sp500_clean), 
                            VIX_Returns = coredata(vix_clean))

# Ensure the VIX and S&P 500 data are aligned by Date
combined_data <- na.omit(combined_data)

# Select the first 6 rows to display in the table
head_data <- head(combined_data, 6) |> select(-Date) |> round(digits = 2)

# Use kable from the knitr package to create a nicely formatted table
kable(head_data, 
      col.names = c("S&P 500 Monthly Returns (%)", "VIX Monthly Returns (%)"),
      caption = "Combined S&P 500 and VIX Monthly Returns",
      align = 'c', # Center align the columns
      table.attr = "style='width:75%; margin:auto;'") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = FALSE, 
                position = "center")

```


## Detailed Analysis of Short-Term Market and VIX Fluctuations

To understand the relationship between short-term market fluctuations and future short-term volatility expectations as represented by the VIX, the analysis began with fitting an ARIMA (AutoRegressive Integrated Moving Average) model to the S&P 500 return data. The ARIMA model is utilized to forecast future values based on past values of a time series and its lags. In this context, the residuals from the ARIMA model—representing unexpected values or shocks in the market—were of particular interest. These residuals signify deviations from the expected trend and cyclical components, effectively isolating the pure shock elements in the market returns, which are crucial for understanding reactionary market behaviors.

### ARIMA Model for Market Returns

The ARIMA model was chosen for its ability to model various types of time series with trends and seasonalities effectively. By analyzing the residuals, we can identify the short-term shocks in market returns, which are not explained by the model's fitted values. These shocks are essential for analyzing how unexpected changes in market returns might influence future market conditions.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-monthly-return-fitted
#| fig-cap: Fitted Values of the Arima Model 

# Fit ARIMA model on S&P 500 monthly returns

arima_sp500 <- auto.arima(sp500_clean)

# Plot the original data and the fitted model

plot(monthly_returns, main="Fitted ARIMA Model", col="blue")
lines(fitted(arima_model), col="red")
legend("topright", legend=c("Actual", "Fitted"), col=c("blue", "red"), lty=1)

```


### Discrete Haar Transform for Decomposition

For both the market return shocks and the VIX data, a Discrete Haar Transform (DHT) was applied to further analyze these time series. The DHT is particularly suited for this analysis because it decomposes the time series data into levels of detail that correspond to different time scales. In our case, the focus is on capturing fluctuations that span up to 2 months, aligning with our definition of short-term fluctuations.We use the 'wavethresh' [@wavethresh] package for this.

The transformation process through DHT breaks down the original time series into a series of approximations and details. The approximations represent the smoothed-out trends, while the details capture the finer nuances of the data at various intervals. For this study, the first level detail coefficients (Detail 1) were extracted, as they represent the highest resolution of short-term fluctuations accessible from the transform, corresponding to a bi-monthly frequency. This choice aligns with our interest in understanding the immediate impacts within a two-month period, considered a critical timeframe for reactive market behaviors.



### Decomposition of Residuals and VIX

Similarly, the VIX data underwent DHT to isolate its short-term components. By focusing on the same level of detail as the market returns, we ensure a consistent basis for comparison. The Detail 1 coefficients from the VIX transformation provide a mirrored view of how volatility expectations, as captured by the VIX, respond in the short term, potentially influenced by the market shocks identified from the S&P 500 residuals.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-decomposed-return-VIX
#| fig-cap: Short Term Fluctuations in Market Shocks and VIX


# Apply Discrete Haar Transform to ARIMA residuals
dht_residuals <- dwt(residuals(arima_sp500), filter = "haar", boundary = "periodic")
details_residuals <- dht_residuals@W

# VIX data analysis using DHT
vix_data_vector <- coredata(vix_clean)
vix_dht <- dwt(vix_data_vector, filter = "haar", boundary = "periodic")
details_vix <- vix_dht@W

# Plotting detailed coefficients of residuals
par(mfrow = c(2, 2))
plot(details_residuals[[1]], type = 'l', 
     main = paste("Detail Residuals at Level", 1),
     xlab = "Coefficients", 
     ylab = "Detail 1 Residuals")
plot(details_vix[[1]], type = 'l', 
     main = paste("Detail VIX at Level", 1),
     xlab = "Coefficients", 
     ylab = "Detail 1 VIX")

```


This detailed decomposition allows us to trace the direct and immediate impacts of market shocks on future volatility expectations, offering insights into the behavioral dynamics of market participants. The analysis of these coefficients in the next section is aimed at exploring whether short-term shocks in the market have a predictive power over short-term shifts in volatility expectations, challenging the conventional wisdom suggested by the efficient market hypothesis.

## Measurement of Short-Term Market Fluctuations and Volatility Expectations

### Quantifying Short-Term Market Fluctuations

Short-term market fluctuations in the S&P 500 index are fundamental indicators of transient market behaviors that can provide insights into investor sentiment and market dynamics. These fluctuations are measured using the monthly returns derived from daily closing prices sourced from Yahoo Finance. The transformation from raw daily prices to monthly returns involves calculating the percentage change in closing prices at the end of each month. This approach smooths out day-to-day volatility, focusing on significant monthly changes which better represent short-term market trends.

The ARIMA (AutoRegressive Integrated Moving Average) model plays a pivotal role in this analysis by fitting these monthly returns to forecast future values and, crucially, to extract the residuals. In econometrics, residuals from such models represent the unexpected component of returns, which can be interpreted as the short-term shocks or the novel information arriving in the market during that month. These residuals are what we define as the "short-term market fluctuations." By focusing on residuals, we directly measure the impact of new information on market behavior, excluding the predictable component based on past data.

### Measuring Future Short-Term Volatility Expectations

Future short-term volatility expectations are captured through the VIX index, often referred to as the market's "fear gauge." The VIX index is designed to measure the stock market's expectation of volatility over the coming 30 days, computed from the option prices of the S&P 500 index. For our analysis, similar to the S&P 500 data, we sourced daily VIX data from Yahoo Finance, which we then processed into monthly averages to align with our market return data.

To specifically measure future expectations as influenced by current market conditions, we applied the Discrete Haar Transform (DHT) to both the residuals of the S&P 500 returns and the VIX data. The DHT helps in decomposing these time series into various levels of detail, reflecting different time scales of volatility. The first level detail coefficients from this transformation represent the highest frequency components available in our monthly data, corresponding to movements within a two-month window. These coefficients capture the essence of rapid changes in volatility and are used to represent the short-term future volatility expectations.

### Methodological Justification and Application

The choice of the ARIMA model for analyzing S&P 500 returns and the application of DHT on both S&P 500 and VIX indices are rooted in financial econometrics, where the objective is to distill raw financial data into actionable insights about underlying market behaviors. This measurement approach ensures that our study accurately reflects the transient phenomena occurring in the financial markets and how they translate into entries in our dataset. By capturing and quantifying these short-term dynamics, we can explore their predictive power on future market conditions.

In summary, our measurement techniques are meticulously designed to transform high-frequency financial market data into a structured format that captures essential dynamics at a monthly frequency. This enables a rigorous examination of the relationship between short-term market shocks and future volatility expectations, providing a robust framework for understanding and forecasting market behavior.

# Model

## Model set-up

In this section, we frame the estimand in terms of a logistic regression model and subsequently use the model to find relationships between the response variable and input variables. We first specify the model and then justify its appropriateness to conduct the analysis. 

### Model Specifications ###

The model used in this analysis is a multiple logistic regression. The variables along with their distributions are given as below:

\begin{align}
y_i|\pi_i &\sim \mbox{Bern}(\pi_i) \\
\text{logit}(\pi_i) &=  \beta_0 \, + \, \beta_1 \times \text{gender}_i 
\, + \, \beta_2 \times \text{education}_i \, +  \, \beta_3 \times \text{race}_i \\
&\quad \, + \, \beta_4 \times \text{economic outlook}_i \,
+ \, \beta_5 \times \text{income change}_i \\
\beta_0 &\sim \text{Normal}(0, 2.5) \\
\beta_1 &\sim \text{Normal}(0, 2.5) \\
\beta_2 &\sim \text{Normal}(0, 2.5) \\
\beta_3 &\sim \text{Normal}(0, 2.5) \\
\beta_4 &\sim \text{Normal}(0, 2.5) \\
\beta_5 &\sim \text{Normal}(0, 2.5)
\end{align} 

where:

1. $y_i$ takes the value 1 if the respondent voted Biden and 0 if they voted Trump
2. $\pi_i$ is the conditional probability of voting for Biden given respondent $i$
3. $\mbox{Bern}(\pi_i)$ is the Bernoulli Distribution with parameter $\pi_i$
4. $logit(x)$ is the logit function that maps $(0, 1)$ to $(-\infty, \infty)$
5. $\text{gender}_i$ is the $i^\text{th}$ respondents gender which takes the value 1 for male and 0 for female
6. $\text{education}_i$  is the $i^\text{th}$ respondents education level which takes values from 1 to 6 depending on the respondents education level
7. $\text{race}_i$ is the $i^\text{th}$ respondents race which takes values from 1 to 7 depending on the respondents race
8. $\text{economic outlook}_i$ is the $i^\text{th}$ respondents view on the economy which takes values on a scale from 1 to 6 depending on the respondents view of the economy
9.  $\text{income change}_i$ is the $i^\text{th}$ respondents change in income which takes values on a scale from 1 to 5 depending on if the respondents income changed for the worse or better
10. $\beta_0$ is the intercept of the logistic regression equation
11. $\beta_1$ is the coefficient of the gender variable
12. $\beta_2$ is the coefficient of the education variable
13. $\beta_3$ is the coefficient of the race variable
14. $\beta_4$ is the coefficient of the economic outlook variable
15. $\beta_5$ is the coefficient of the income change variable

The above model will be used later in the results section to find the impact of the input variables on voter preferences. For now, we justify certain characteristics of the model that capture certain real life aspects of the analysis. 

### Model Justification ###

Our goal is to find the effect that certain voter specific characteristics and non-partisan views have on a voter voting for Joe Biden. For that, we consider the framework of the model by considering the response variable, the input variables, and how they come together within the model structure.

#### Response Variable ####

Our variable of interest is voter preferences for Joe Biden. We assume that voting preference for Biden is a random variable and has a $\mbox{Bern}(\pi_i)$ distribution with $\pi_i$ being the 'likelihood of voting for Biden'. We choose specifically the random variable characterization because voting outcomes aren't deterministic but rather phenomenon subject to chance. This also aligns with the voting patterns we observe in the real world. 

We estimate the 'likelihood of voting for Biden' by the using the conditional probability of voting for Biden given a certain respondent. 

#### Input Variables ####

The voter specific characteristics that we consider are:

1. Gender
2. Race
3. Education
4. Change in Income

The non-partisan view of the respondents that we consider is:

1. Economic Outlook 

We assume for this analysis that all 5 variables have Normal prior distributions as based of our prior information, we can't conclude any specific form for any of the variables' distributions. All we can conclude is that the extreme values taken by each variable are less likely than their intermediate values and hence a suitable distribution to describe flat tails with a peak in the center is the normal distribution.

#### Model Structure #####

We construct a model that gives the conditional probability of voting for Biden from only the above 5 input variables. Probability is a measure taking values from $[0,1]$ which is why we use the logit function to transform the conditional probability to the unbounded real line. The logit function is appropriate in this context because it preserves the ordering of probabilities while expanding them to the entire real line. 

Once we have the $logit(\pi_i)$, we define a linear relationship between this quantity and the 5 categorical variables. Taking the inverse of the logit function gives probability as a function of the 5 categorical variables. We then use this model to study the impact of the 5 categorical variables on voting for Biden. 

\begin{align}
\text{logit}(\pi_i) &=  \beta_0 \, + \, \beta_1 \times \text{gender}_i
\, + \, \beta_2 \times \text{education}_i \, +  \, \beta_3 \times \text{race}_i \\
&\quad \, + \, \beta_4 \times \text{economic outlook}_i
\, + \, \beta_5 \times \text{income change}_i \\
\pi_i &= \text{logit}^\text{-1} \text{(} \beta_0 \, + \, \beta_1 \times \mbox{gender}_i \,
+ \, \beta_2 \times \text{education}_i \, +  \, \beta_3 \times \text{race}_i \\
&\quad \, + \, \beta_4 \times \text{economic outlook}_i \,
+ \, \beta_5 \times \text{income change}_i \text{)}
\end{align}

#### Parameter Estimation ####

The model has 6 parameters  $\beta_0$, $\beta_1$, $\beta_2$, $\beta_3$, $\beta_4$, $\beta_5$ which need to be estimated from the cleaned_data dataset. These parameters are estimated by defining the model within the stan_glm() function from the [@Rpackage_rstanarm] package. The function uses Markov Chain Monte Carlo (MCMC) algorithms to create a chain of samples from the posterior distribution of each coefficient. This not only gives us point estimates, but also the posterior distribution of values for each coefficient. 

The posterior distribution contains information about all of the coefficients moments, giving us estimates for not only the central values, but also the uncertainty associated with each estimate. Using the central moments as well as the 95% credibility intervals for the coefficients, helps us determine whether the input variables have an effect on voting preferences. 


# Results

This section looks at the results of the regression performed using the Model from the Model section and the cleaned data set. We first look at the results of our regression analysis, which includes, the parameter estimates and their credibility intervals. Subsequently, we explore select variables of interest and their effects on the probability of voting for Biden. Finally, we assess model fit using the $R^2$ metric and compare our final model's $R^2$ to the same model with certain terms dropped in order to assess the explanatory power of our model.  

## Regression Results

@tbl-coefficient-estimates contains the parameter estimates for the regression analysis. The regression analysis was done on a random sample of 10,000 observations out of 39,203 in order to speed up computational time while preserving the properties of the dataset. The first column of the @tbl-coefficient-estimates contains the parameter name. Since all the input variables are categorical variables, the regression creates indicator terms for each value the variable takes and estimates a unique coefficient based of that. 

The second column contains the actual parameter point estimates. These estimates are based of the stan_glm() function from the [@Rpackage_rstanarm] package. The third column gives the standard error while the 4th and 5th columns give the lower and upper limit for the credibility intervals. The table was created from the model using the tidy() function from the [@Rpackage_broom_mixed] package, while the 'kable()' and 'kable_styling()' functions were used from [@Rpackage_knitr] and [@Rpackage_kableExtra]. 

@fig-coefficient-estimates creates a plot of @tbl-coefficient-estimates with the estimate and standard errors shown. The plot shows the credibility intervals for each parameter.


```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-coefficient-estimates
#| tbl-cap: Parameter Estimates of the Multiple Logistic Regression

library(broom.mixed)
library(knitr)
library(kableExtra)

coeff <- tidy(voter_outcomes_model, conf.int = TRUE)
cleaned_coeff <- kable(coeff, col.names = c(
                        "Parameter", "Estimate", "Standard Error",
                        "Credibility Interval Lower Limit", 
                        "Credibility Interval Upper Limit"),
                         digits = 2)
cleaned_coeff
```

```{r}
#| echo: false
#| eval: true
#| warning: false
#| label: fig-coefficient-estimates
#| fig-cap: Parameter Estimates and Confidence Intervals

modelplot(voter_outcomes_model, conf_level = 0.9) +
  labs(x = "90 per cent credibility interval")
```

From the table and the plot, we observe that the gender variable has a negative coefficient estimate, indicating that given the respondent is male holding all else fixed, they have a lower probability of voting Biden compared to Trump. The coefficient estimates for education are all positive indicating that a respondent with a high school degree or higher has a higher probability of voting for Biden over Trump. The coefficient estimates for Race are positive for all categories except 'two or more races' and 'native american' indicating that given a respondent is either 'Black', 'Asian', 'Middle Eastern' or 'Hispanic', they have a higher probability of favoring Biden.

The economic variables which are 'economic outlook' and 'income change' have distinct patterns compared to the demographic variables which are 'Gender', 'Education', and 'Race'. For one, the economic outlook coefficients, given in @tbl-coefficient-estimates are all positive but much larger in magnitude than any of the other variable coefficients. This indicates that given a respondent who's outlook on the economy is between neutral and poor, we find that the respondent has a very high probability of favoring Biden over Trump.

The same holds for change in income as most of the income coefficients are positive but negligible. This indicates that the respondents preferences based of income change are not very significant.

### Model Fit ###

We now assess the fit of the model to the data. The statistic that validates model fit in this situation is $R^2$. We compare our models $R^2$ with the $R^2$ of the model:

1. Taking only demographic variables (Gender, Race, Education)
2. Taking only economic variables (Economic Outlook, Income Change)

These 2 different variants of the model are taken to show the improvement that our final model makes over both of them. We consider $R^2$ as a measure of model fit as it's a measure of the variation within the data explained by the model. The $R^2$ of all 3 models are given in @tbl-R-squared-estimates. The first column lists the parameters along with different statistical measures, including $R^2$. The second column gives the gives the coefficient estimates and $R^2$ of the main model. The third column gives the coefficient estimates and $R^2$ of the model taking only demographic variables as the input variables. The third column gives the same but for economic variables. 

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: tbl-R-squared-estimates
#| tbl-cap: Models Coefficients and R-Squared Estimates

library(modelsummary)

model_summary <- modelsummary( 
  list (
    "Main Model" = voter_outcomes_model, 
    "Economic Only" = voter_outcomes_demographic_only_model,
    "Demographic Only" = voter_outcomes_economic_only_model
      )
    )
model_summary
```

@tbl-R-squared-estimates shows that the $R^2$ for the main model is 0.458 while the $R^2$ for the demographic only model and economic only model is 0.104 and 0.383 respectively. This shows that the demographic only model's fit is significantly less than our final model's, showing that considering economic variables in the analysis significantly boosted model fit. 

When we look at the model with only economic variables, we find that a good amount of variation in model is explained by these 2 variables, evidenced by the comparatively high $R^2$ of 0.383. The model though lacks some interpretability and can be improved by considering the demographic variables on top. This leads us back to our main model, with the highest $R^2$ between the 3.

# Discussion

## Demographic Characteristics Effect on Voter Preference

In this paper we examined how race, education, gender, economic outlook and change in income impacted voters’ preferences for Biden over Trump in the 2020 US presidential election. Our analysis showed that Biden voters thought the nation's economy got much worse. Furthermore, voters that supported Biden were female, Black, Hispanic, and Asian. They possessed a minimum of a four-year college degree or higher. Finally, they were white and non-white college graduates as well as non-white without a four-year college degree. Overall, Biden voters were women, more racially diverse with greater education attainment.

Our data and other research we conducted on education as well as education by race was consistent. There were a few discrepancies in the data we analyzed and previous research with regards to gender and race. 

First, voters’ gender on voting preferences were contradictory. Our data demonstrated a greater percent of women voted for Biden whereas [@citeIgielnik] points out that Biden received more votes from men. A plausible explanation for this difference is a small difference in gender preferences. Further research should be done as well as observing previous election results to determine whether voters' gender really is a reliable predictor of voting preferences in US elections.

Secondly, our data on race was congruent with Whites, Blacks and Asians. Our data indicated Hispanics voted for Biden however other research suggests Hispanics were more inclined to vote for Trump. Another incongruence involves the 'other' race category in that our data signifies the other group voted Trump however research alludes this group as being Biden supporters. A possible reason for this inconsistency is that in our data, the other category consists of only three percent of electorates and Hispanics ten percent. Having such a small representation of these groups increases the likelihood of discrepancies in results.

## Economic Characteristics Effect on Voter Preference

Perhaps the strongest effect on voting preference, was through the economic outlook variable. The coefficients were on orders of magnitude 2-3 times larger than the coefficients for the other variables. This indicates the prevalence of a poor economy sentiment among majority of the voter base. This also shows that the prevelance of a weak economy during the pandemic influenced the political preferences of voters and shifted a large amount of the populace in favor of Biden.

This, coupled with the fact that a majority of voters who saw their incomes reduce favored Biden, alludes to the possibility that a weak economy during election season prompted an anti-incumbency sentiment amongst many voters, influencing them to vote for Biden.

The effect of both these variables on voting preference for Biden is interesting because, for one, they had the strongest combined effect, and also are characteristics which aren't easily predictable based of a voters demographics. Also, when we ran our regression with only these 2 variables, they had better predictive power than the model with only the demographic variables involved. Therefore, economic characteristics may be the key to understanding voter preferences, and combined with demographic characteristics, tell a compelling story of voter outcomes.
 
## Weaknesses 

A major weakness of our analysis is assuming a linear model for voting preferences as a function of demographic and economic characteristics. Assuming linearity hids a lot about the mechanism through which these characteristics influence voting outcomes. Though model interpretability is increased, we lose any sound understanding of the mechanism underlying the phenomenon that we aim to model.

## Next steps 
 
Despite taking place during COVID-19, the 2020 election showed the highest turnout in the 21st century, where 66.8% of eligible voters voted. This is up from 60.1% in the previous election in 2016, most likely due to the addition of nontraditional methods of voting such as early voting and mailing ballots to ensure adequate voter turnout [@citeFabina]. A potential probe is looking at the demographics behind those that voted by mail, absentee, early or on the traditional day. For example, [@citeIgielnik] states Hispanic and whites voted by mail whereas blacks voted early in person.

A possible area of additional research could question the correlation between voters’ race and the race of the candidate they vote for. For instance, investigating the different demographics that voted for Biden, a white male candidate compared to his running mate, Kamala Harris a black female candidate. Additionally, white respondents who showed high levels of radical resentment were more likely to vote for Joe Biden as opposed to Kamala Harris. Not only was Harris at a disadvantage because of her race but also because of her gender. [@citeNelson] states that “hostile sexism” may play a role in Harris’s success in the election (2021). The combination of these two variables could lead to interesting findings.

Additional research could look in the direction of age as a predictor in the US election. [@citeLees], with the use of 1000 research participants through a multi-step design, finds voters aged 49 and younger were in support of Biden whereas the older population favored Trump, however, not by a significant amount. Another trend portrayed the insignificance of candidates age on voter preference. Young adults aged 30 and below indicated that voting for a candidate that is over the age of 70 did not change their voting decision. At the time of the election, both Biden and trump were over the age of 70 [@citeLees]. Thus further research is necessary to determine the reliable relationship between age and voting preferences.


\newpage

# References


